---
title: "overview"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{overview}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---
  
```{r, include = FALSE}
knitr::opts_chunk$set(
collapse = TRUE,
comment = "#>"
)
```

```{r setup}
library(MultiOpt)
```

# Basics of simulated annealing

# understanding this package: what it does and doesnt do

# A simple example

## Loading data

For now we will generate some data to look at. This data is weakly positively correlated.
```{r}

n <- 100 # how many "individuals" we want in our sample

set.seed(12345)
x = rnorm(n = n, mean = 120, sd = 2)
y = x * 4 + rnorm(n = n, mean = 0, sd = 5)
z = y * 2 + rnorm(n = n, mean = 0, sd = 5)

dat = data.frame(x = x, y = y, z = z)

head(dat)
```


## Formatting data

### "trait" data
Expected formatting for trait data is a list of matrices. Here we are just working with vectors, but pairwise matrices of distances (genetic or otherwise) can also but added to the list.

The package refers to provided data as "trait" data, but this is in the broad sense and can include phenotypic or genotypic data, so long as it can be put into a matrix. 
```{r}
trait_list = list(
  x = as.matrix(dat$x),
  y = as.matrix(dat$y),
  z = as.matrix(dat$z)
)
```


This list would work, however, as soon as you try and run the simulation you would notice a warning about the scales of the traits being off. This is an important consideration for your simulation. If the scales of your traits are considerably different, the trait with the larger values would be "favored" in the simulation if `p_depends_delta = TRUE`, as the larger differences in values between samples are larger and thus more important to the algorithm. Thus, its a good idea to transform your data before running simulated annealing. Fortunately, we have a function to do so. 

```{r}
trait_list_scaled <- scale_traits(trait_list)
```


### Measures
Now we need to specify what out exact goals are for each of these traits. Some basic goals would be minimizing or maximizing a trait (e.g., minimizing inbreeding vs maximizing genetic diversity), or even maximizing variance (e.g., preserving phenotypes preventing selection when subsetting). There are quite a few options, see `multiopt_sa()` for more details. INSERT TABLE WITH DESCRIPTIONS HERE??? We can pick the direction of the measure in the arguments parameter, described next.

Again, this information needs to be in a list, with each list element being the function we want to use to measure success. For simplicity, we'll just go with summarizing the measures with a mean of our trait data, which are vectors, and thus why we will pick `weighted_mean_of_vector`. Weighting is indicating the mean is modified by how many times an individiual is chosen. 
```{r}
measure_list = list(
  x = weighted_mean_of_vector,
  y = weighted_mean_of_vector,
  z = weighted_mean_of_vector
)
```

### Measure arguments
Some arguments require specialized parameter specification; here is where you can add those details. The arguments are also where you can specify what direction of success you are interested in (minimizing or maximizing, specified by direction = 1 or -1). Again, this needs to be a list. All trait names need to match between the trait, measures, and argument lists.

Let's choose to maximize the first two traits (`x`, and `y`), but minimize `z`. Since we are just doing a simple mean, we dont need to supply any other arguments
```{r}
args_list = list(
  x = NULL, # dont need any special arguments here
  y = NULL,
  z = list(direction = -1) # -1 here indicates we want to get the minimum value of the measure
)
```


## A basic analysis

```{r}
# how many "individuals" or samples we want in the final population
n_t = 50 

# maximum number of times an individual can be chosen in the population. 
# Here we are selecting an individual a maximum time of 1. 
# You could select an individual multiple times for example when taking cuttings or collecting seeds from a selfed individual.
weights_max = 1 

# initial_weights = sample(c(rep(1, n_t), rep(0, n-n_t))) # if you have an idea of what individuals are going to be best you can start the simulation off with those. There we are just picking the first `n_t` individuals. This argument isnt needed if specifying n_t.


```

And with that we can start a simulation:
```{r}
out <- multiopt_sa(
  trait_list = trait_list_scaled,
  measure_list = measure_list,
  measure_args_list = args_list,
  n_t = n_t, 
  weights_max = weights_max,
  verbose = F
)
```


### Take a look at the results

```{r}
plot(out$chain$values[,1], type = "l") # trait x
plot(out$chain$values[,2], type = "l") # trait y
plot(out$chain$values[,3], type = "l") # trait z
```

What we want to see an lots of bouncing around in the beginning, indicating the algorithm was exploring. As the temperature cools, we want to see a gradual increase in value, with the final iterations being fairly stable. With three trait dimentions the trait relationships can get a bit convoluted. Here we can see very nice simulation results of our "chain" (what the algorithm decided as the temperature cooled) for two of the traits (`x` and `y`), but the opposite of what we would expect for the last trait (`z`). If we think back to how we generated our data, we made all the traits weakly positively correlated, then chose to maximize two of them and minimize the last one. So it was easy for the algorithm to maximize two traits (since they were positively correlated), but since the last trait was going in the opposite direction, it was not prioritized (as two traits being improved is better one trait improving). So these results make sense in the setup we provided. 


We can also back transform the output to make the results a bit more interpretable.
```{r}
out_unscaled <- unscale_multiopt(trait_list = trait_list, multiopt_output = out)

plot(out_unscaled$chain$values[,1], type = "l") # trait x
plot(out_unscaled$chain$values[,2], type = "l") # trait y
plot(out_unscaled$chain$values[,3], type = "l") # trait z
```



## Understanding what parameter settings are doing
One of the limitations with simulated annealing is that you will need to modify 
the starting parameters for each new combination of data and objective, as the
analysis can be quite sensitive to each of the settings

temperature (min and max)
iterations (max_steps)




## A more complicated analysis
### Randomizations
Simulated annealing, by definition, is a stochastic optimizer. As such, it is almost certain that
results of each simulated annealing run will result in slightly different final results. 
As such, it would be good practice to run the process a few times to confirm the general 
range of expected values. Replication will also allow you to see which individuals are
more frequently selected for the final populaiton mix, allowing you to prioritize which 
individuals to select for your goals.


MultiOpt has a wrapper for `multiopt_sa` that allows for replication of simulated annealing
with a simplified output of the results, making it easier to get a nice list of
replicated results: `rand_multiopt``. Parameters are virtually the same between 
`multiopt_sa` and `rand_multiopt` since they are doing the same task. The only additional
information `rand_multiopt` needs is how many times you want to repeat the analysis (`n_runs`),
and if you want to run the repeats in parallel (`parallel`). Note that it probably
isn't necessary to run analyses in parallel unless you are doing a lot of replicates or very long
simulations.



Using the same data and settings we generated earlier, the setup is very similar:
```{r}
# arguments for simulated annealing need to go into a named list
sa_args = list(
  trait_list = trait_list_scaled,
  measure_list = measure_list,
  measure_args_list = args_list,
  n_t = n_t, 
  weights_max = weights_max,
  max_t = 2,
  max_steps = 5000, # this should be increased when running an actual analysis
  nda = T,
  nd_samples = 500
)

multi_out = rand_multiopt(n_runs = 5, multiopt_args = sa_args, parallel = F)

str(multi_out)

```


To run in parallel you simply need to set up the session and specify `parallel = T`:
```{r}
library(future)

future::plan(multisession, workers = 4) # multisession indicates this is run on current machine
multi_out = rand_multiopt(n_runs = 5, multiopt_args = sa_args, parallel = T)

str(multi_out)

```


And let's look at the results for the first two objectives (traits):
```{r}
plot(multi_out$measure_summaries)
```

As you can see, results are similar for each iteration but not quite the same. 



### Exploring the Pareto Front further

While the final solutions from simulated annealing are informative, we are also 
interested in trade offs between our objectives are, or the Pareto Front. 
The Pareto front are the solutions found during the simulated annealing 
run that are "non-dominated", or the best choices given multiple conflicting goals. 
A solution (an option in the simulated annealing chain) is non-dominated if you cannot
improve an objective without sacrificing performance in at least one other objective.

In short, the Pareto Front shows us what the best-case solutions are when there are trade-offs.
The collection of these non-dominated solutions are the Pareto Front, or the "archive."
An archive can be generated by both `multiopt_sa` and `rand_multiopt` with the setting `nda`.
all the archive values generated by each run in `rand_multiopt` are combined into a single
archive, which is helpful to give a more complete idea of what the tradeoffs are. 
If you wish to try and get more values for your archive you can use `expplor_pareto`,
which uses the already generated Pareto Front as a starting point to keep looking 
for non-dominated solutions.

```{r}

pareto_out = explore_pareto(
  multi_out$archive,
  trait_list = trait_list_scaled,
  measure_list = measure_list,
  measure_args_list = args_list,
  n_t = 50,
  weights_max = weights_max,
  max_t = 2,
  nd_samples = 200,
  verbose = F
)

plot(pareto_out$archive_summary)
```


#### comparing multi-objective to single objective
In addition to the Pareto Front, it can be helpful to look at the trade offs
when looking at a single objective vs multi-objective goals. If there is a small
tradeoff between two objectives, you don't really need to even measure both of them.


`singleopt_context` runs single-objective simulated annealing for each of the provided
traits and calculates the resulting measures for each of the other traits. This analysis,
in combination with the Pareto Front gives a fairly complete view of the trade-offs
between objectives.


Here is a full analysis in one swoop:
```{r}
n <- 500 # how many individuals to generate trait data for

# strongly + correlated
set.seed(12345)
x = rnorm(n = n, mean = 120, sd = 2)
y = x * 4 + rnorm(n = n, mean = 0, sd = 5)
z = y * 2 + rnorm(n = n, mean = 0, sd = 5)

dat_strong_p = data.frame(x = x, y = y, z = z)

trait_list = list(
  x = as.matrix(dat_strong_p$x),
  y = as.matrix(dat_strong_p$y),
  z = as.matrix(dat_strong_p$z)
)


trait_list_scaled <- scale_traits(trait_list) # scale trait data

measure_list = list(
  x = weighted_mean_of_vector,
  y = weighted_mean_of_vector,
  z = weighted_mean_of_vector
)

args_list = list(
  x = NULL,
  y = list(direction = -1),
  z = NULL
)

n_t = 50
weights_max = 1

sa_args = list(
  trait_list = trait_list_scaled,
  measure_list = measure_list,
  measure_args_list = args_list,
  n_t = 50, weights_max = 1,
  max_t = 2,
  max_steps = 5000, # this is quite short, but will give a general idea
  nda = T,
  nd_samples = 200,
  verbose = F

)

multi_out = rand_multiopt(n_runs = 10, multiopt_args = sa_args)

# run multiple rounds of SA
multi_out_unscaled = unscale_rand_multiopt(trait_list, multi_out)

# explore pareto more
pareto_out = explore_pareto(
  multi_out$archive,
  trait_list = trait_list_scaled,
  measure_list = measure_list,
  measure_args_list = args_list,
  n_t = 50,
  weights_max = 1,
  max_t = 2,
  max_steps = 5000,
  nd_samples = 500,
  verbose = F
)

pareto_unscaled = unscale_archive(trait_list, pareto_out)

# compare to single objective
single_out <-
  singleopt_context(
    trait_list = trait_list_scaled,
    measure_list = measure_list,
    measure_args_list = args_list,
    n_t = n_t,
    n_runs = 10,
    verbose = F
  )

single_out_unscaled = unscale_singleopt(trait_list, single_out)
```


MultiOpt has a couple helper functions to make some quick summary plots. Here
we can look at the Pareto Front using plot_pareto().

```{r, fig.width=10}
# plot
library(dplyr)
library(ggplot2)

# pareto front
plot_pareto(
  archive_list = pareto_unscaled,
  multi_list = multi_out_unscaled, # optional
  single_list = single_out_unscaled # optional
)

```

We can also plot which individuals were selected in each iteration.
```{r, fig.width=10}
plot_selection(multi_out_unscaled$individs_selected, trait_list = trait_list)

# and for the single objective optimization results, we can look at who was 
# selected in the context of the other traits
plot_selection_single(single_out_unscaled$individs_selected, trait_list = trait_list)
```



And there is even a function to help make pretty chain plots all in one go.
```{r, fig.width=10}
plot_chain(out_unscaled$chain)
```



We can also take a look at a 3d plot of pareto front using Plotly.
```{r}
library(plotly)

test_single_df <- dplyr::bind_rows(single_out_unscaled$measure_summaries, .id = "trait") %>% # single objectives
  dplyr::as_tibble()

plot_ly(x = ~x, y = ~y, z = ~z) %>%
  add_markers(data = test_single_df, color = ~trait) %>%
  add_markers(data = as.data.frame(multi_out_unscaled$measure_summaries), color = "multi") %>%
  add_markers(data = as.data.frame(pareto_unscaled$archive_summary), color = "Pareto")

```






